\name{}: machine learning over interaction networks and ontology-based background knowledge predicts drug--target interactions

### 2021-07-01 paper/main.tex
.

% With the rapidly growing field of machine learning approaches and
% their application to bioscientifical problems in the realm of
% bioinformatics, different kinds of data, such as long DNA sequences
% could be utilized for feature generation, while rapid advances were
% made. 
Identifying drug--target interactions (DTIs) is one of the initial steps in drug discovery and screening for disease treatment and thus helps to mitigate the the costs and risks of drug development.  Further, it forms a generalization of both drug repurposing and repositioning, finding novel functions for already approved and extensively analyzed drugs with well explored properties. Inferring and deriving from those existing properties and interactions can help to analyze and identify novel drugs for known targets, leading to desirable therapeutic effects. Moreover, while \textit{in vitro} DTI prediction requires the availability of all substances, computational, \textit{in silico} DTI predictors interpret and anticipate millions of interactions within short time.

@@

Both bottom-up and top-down approaches to drug--target interaction
prediction bear some advantages and limitations. Generally, bottom-up
methods face the challenge to predict whether a chemical structure binds to a
protein given their molecular properties; whether two entities
interact depends not only on the molecular structure of the entities
(where binding sites and molecular forces need to be determined for
accurate prediction) but also properties namely in which celltypes and anatomical structures a protein is expressed. Top-down
methods use information for DTI prediction such as side-effect
similarity \cite{Campillos2008} that is largely complementary to
knowledge gained from molecular properties. While methods that rely on
molecular information are directly predicting whether two molecules
could interact, top-down methods base on more indirect means and 
infer a DTI from observable effects resulting from the interaction.

@@

We further investigate the influence of the more specific interaction types \textbf{inhibition} and \textit{activation} within STITCH database on the performance of \name{}. The results are summarized in Supplementary table 3, where we find no difference in comparison to the complete dataset.

>>>>

% With the rapidly growing field of machine learning approaches and
% their application to bioscientifical problems in the realm of
% bioinformatics, different kinds of data, such as long DNA sequences
% could be utilized for feature generation, while rapid advances were
% made. 
Identifying drug--target interactions (DTIs) is a crucial step in drug
discovery; finding novel DTIs for approved drugs can be used for drug
repurposing, either by finding new drugs for a known target or finding
a drug for a novel target involved in a disease process.  Inferring
the interactions between drugs and their targets can help to analyze
and identify potential desired or adverse drug effects as well as
desirable therapeutic effects. While \textit{in vitro} DTI prediction
is time consuming, computational \textit{in silico} DTI predictors can
screen for millions of interactions within a short time.  Determining
DTIs computationally can therefore help to mitigate the costs and
risks of drug development.

@@

Both bottom-up and top-down approaches to drug--target interaction
prediction bear some advantages and limitations. Generally, bottom-up
methods face the challenge to predict whether a chemical structure
binds to a protein given their molecular properties; whether two
entities interact depends not only on the molecular structure of the
entities (where binding sites and molecular forces need to be
determined for accurate prediction) but also properties such as in
which celltypes and anatomical structures a protein is
expressed. Top-down methods use information about physiological
effects of drugs for DTI prediction, such as side-effect similarity
\cite{Campillos2008}, that is largely complementary to knowledge
gained from molecular properties. While methods that rely on molecular
information are directly predicting whether two molecules could
interact, top-down methods base on more indirect means and infer a DTI
from observable effects resulting from the interaction.

@@

We further investigate the performance on the specific interaction
types {\em inhibition} and {\em activation} that are available within
the STITCH database. The results are summarized in Supplementary Table
3; we find no difference between the performance on specific
interaction types and  the complete dataset where we do not separate
types of interaction.
